A Philosophy of (Inductive) Testing

### 2024-02-20 mss.tex
Semi-final version (missing page numbers).

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              Manuscript submitted for review.\\
              % {\it Urbit Systems Technical Journal} I:1 (2024):  1–3. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{What Should I Test?}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{What Should I Test?}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

Under the covers the Obelisk engine calls out to the urQL parser to create AST commands from the user-created scripts. The \texttt{AS OF} clause overrides the time passed in the bowl. Finally we weld together two results and check them against what we expect. The first check, expect-eq, is of the metadata returned by the last command. The second check is the final database state after the evolution.

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} and \texttt{false} \citep{Yarvin2006}, meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. When speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing, $\textrm{Succ}(0)$, gives us the first thing.  Applying $\textrm{Succ}$ to the first thing, $\textrm{Succ}(\textrm{Succ}(0))$ gives us the second thing, and so on.

@@

What does induction have to do with testing, however?  For inductive types—and simple functions over inductive types—the software author only has to prove, or test, two cases:  the case for zero and the case for the successor value after zero.  For example, Listing~\ref{lst:sequent} depicts tests for a gate over an inductive type (\texttt{list}) in \texttt{/lib/seq} (\citet{FoxSequent}, currently distributed via \texttt{\%yard} \citep{UrbitYard}).

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \texttt{\%quiz} is a well-documented Hoon implementation \citep{Hjort2023}.  Once again, this solution requires some deep thinking about the code and its specification.

>>>>

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} I:1 (2024):  20–23. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{A Philosophy of (Inductive) Testing}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{A Philosophy of (Inductive) Testing}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

@@

Under the hood, the Obelisk engine calls out to the urQL parser to create AST commands from the user-created scripts. The \texttt{AS OF} clause overrides the time passed in the bowl. Finally we weld together two results and check them against what we expect. The first check, expect-eq, is of the metadata returned by the last command. The second check is the final database state after the evolution.

@@

Peano arithmetic is a better lay programmer's introduction to proof by induction. There exists (or perhaps \emph{does not exist}) a special concept, zero. Playing fast and loose with classical logic, “ex nihil sequitur quodlibet”, from nothing (more commonly a falsehood, or a contradiction) follows everything (i.e. anything).\footnote{The “principle of explosion” rigorously follows from use of disjunctive syllogisms.  Here, we jocosely indicate that by proceeding from zero inductively we can demonstrate desired properties of testing.} Zero is not only an integer; rather, in terms of Nock-based programming, \lstinline[style=inlinecode]{~} is not only the beginning of counting,\footnote{Yarvin opted to invert \texttt{true} (as 0) and \texttt{false} (as 1) \citep{Yarvin2006}, meaning that Nock loobeans do not align inductively with other inductive types.} but the nothing of every inductive type, most importantly trees and lists. When speaking to programmers, Peano tells us there is a universal nothing \lstinline[style=inlinecode]{~} and there is a function called successor $\textrm{Succ}$, which produces some next thing from a previous thing (or lack of thing). Applying $\textrm{Succ}$ to nothing, $\textrm{Succ}(0)$, gives us the first thing.  Applying $\textrm{Succ}$ to the first thing, $\textrm{Succ}(\textrm{Succ}(0))$ gives us the second thing, and so on.

@@

What does induction have to do with testing, however?  For inductive types—and simple functions over inductive types—the software author only has to prove, or test, two cases:  the case for zero and the case for the successor value after zero.  For example, Listing~\ref{lst:sequent} depicts tests for a gate over an inductive type (\texttt{list}) in \texttt{/lib/seq} (\citetrepo{FoxSequent}, currently distributed via \texttt{\%yard} (\citetrepo{UrbitYard})).

@@

An automated testing solution called property-based testing may be applicable in these cases. The idea of property-based testing is to develop and instantiate invariant properties of the code and let software generate random inputs. The software runs the random inputs and tests the outputs against the invariant properties. An architecture for this approach was originally developed for Haskell and has since ported to many other languages, including Hoon.  \texttt{\%quiz} is a well-documented Hoon implementation \citetrepo{Hjort2023}.  Once again, this solution requires some deep thinking about the code and its specification.
